SUPPLEMENT to \\ Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-05-22 doc/mc-supplement.tex
Minor edit
Final edits for REB

Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
delete the input clauses.  Consider the first of these, deleting
input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
that there is no total assignment $\assign$ that falsifies this clause but assigns $\assign(s_{10}) = 1$.
The proof proceeds by first assuming that the clause is false, requiring
$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consist of unit
clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
clauses that cause unit propagation.  Hint clauses 8 and 10 force the
assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 force
$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
via clause 22 to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
completing the RUP proof.  The deletion hints for the other input
clauses follow similar patterns---they work from the bottom nodes of
the POG upward, showing that any total assignment that falsifies the clause
must assign $\assign(s_{10}) = 0$.

@@

Our experiments are designed to evaluate the following:
\begin{itemize}
\item
  whether our toolchain can handle challenging benchmark problems,
\item
the effectiveness of some design choices and optimizations,
\item
the comparative performance of the prototype and verified checkers, and
\item
how the performance of our toolchain compares to that of the MICE model counter verifier.
\end{itemize}

@@

Our default configuration for the proof generator used the structural
method for the forward implication proof, with the two optimizations
of literal grouping and lemmas.
Running with a time limit of 4000 seconds\footnote{All measured times
in this document are actual elapsed times, not CPU times.}, \dfour{}
was able to generate dec-DNNF representations for 124 of these, and it timed out on the
rest.  Our proof generator was able to convert all of the generated
dec-DNNF graphs into POGs and determine how many defining clauses they
would generate.  A POG operation with $k$ arguments requires $k+1$
defining clauses, and so the total number of defining clauses in POG $P$ equals
the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$.
The number of defining clauses ranged from 304 to 2,761,457,765 with a median of 848,784.

@@

Figure~\ref{fig:d4:cpog} summarizes our results in terms of the time
required by \dfour{} (X axis) versus the sum of the times for the
proof generator and checker (Y axis).
We were able to complete a full validation for 108 of the 124
benchmarks, with times ranging from fractions of a second to 13,144
seconds, with a median of 71.6 seconds.

Relative to the runtime for \dfour{}, two problems ran faster with the
generator and checker.  These were for problems having small numbers
of models (relative to the number of variables), and so most of the
time spent by both \dfour{} and our proof generator was in running a
CDCL solver to search the very sparse solution space.  \cadical{}
generally outperforms the miniSAT-based solver used by \dfour{}.  At
the other extreme, one problem that required only 19 seconds for
\dfour{} required 8657 seconds to generate a proof and 45 seconds to
check it.  This benchmark appears to stem from weak performance
by our implementation of BCP\@.  Overall the ratios
between the combined generation plus checking times versus the time
for \dfour{} had a harmonic mean of 5.5.

@@

To assess the relative merits of the two approaches to forward
implication proof generation, we ran the proof generator in monolithic
mode on the 92 problems for which the combination of structural proof generation
and checking required at most 3000 seconds and then kept only those
measurements for which at least one of the two approaches had combined
times (generation plus checking) of at most 1000 seconds.  There were
a total of 83 such problems.  Figures~\ref{fig:monolithic:time} and
\ref{fig:monolithic:clauses} show comparisons between the two
approaches in terms of time and total clauses.

Examining Figure~\ref{fig:monolithic:time}, we can see that the
monolithic approach ran faster than the structural approach for 23 of
the problems, including a case where the generator and checker
required only one second, versus 71 seconds for the structural
approach.  On the other hand, the monolithic approach timed out for 13
of the cases and was slower for the other 37.  In general, we can see
the monolithic approach faring worse on larger problems.

@@

As the figures illustrated, each optimization on its own can shorten
the proofs, sometimes substantially, and the two combine to have even
greater effect.  Of the 80 benchmark problems evaluated, and comparing
with both optimizations disabled, 47 had the number of proof clauses
reduced by a factor of 2 by using lemmas, 14 by using literal merging,
and 60 by enabling both optimizations.  In the extreme cases, a lack
of lemmas caused one proof to grow by a factor of 52.5, while a lack
of literal grouping caused another proof to grow by a factor of 39.6.

@@

One strength of monolithic proof generation is that it makes no
assumptions about how the compiled representation.  A complete
proof-generating SAT solver could, in principle, always generate the
forward implication proof, as long as forward implication holds.  To
test this hypothesis, we ran our proof framework with preprocessing
enabled on the same 83 benchmarks as in
Section~\ref{app:experiment:monolithic}.  We measured the time for two
entire toolchains: one with compilation, proof generation, and
checking, and the other starting with preprocessing and then having
the same steps as the other.  The results are summarized in
Figures~\ref{fig:preprocess:time} and \ref{fig:preprocess:clauses}.
Setting a time limit of 1000 seconds, 70 of the formulas completed
without preprocessing, while 69 did so with preprocessing.  The one
that timed out with preprocessing did so due to the excessive time
required for preprocessing.  Of the 69, 53 ran faster with
preprocessing, while 15 ran slower.  The sizes of the generated 
CPOG proofs were more variable: 31 had fewer clauses with preprocessing, 21 had more, and 17 were identical.

@@

This section provides more insights into how our proof framework (POG)
compares with other work in certified knowledge compilation and model counting.

@@

To compare the performance of our toolchain versus that of
MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
generation, proof checking, and ring evaluation in the CPOG toolchain compares to the time to
run their proof generator \progname{nnf2trace} and checker
\progname{sharptrace}.  We add the time for ring evaluation on our
side, since their checker provides a count, although only an
unweighted one.  We used the prototype versions of the checker and ring evaluator.
We ran their tools on the 92 benchmark problems for
which our toolchain requires at most 3000 seconds and retained
the 84 data points for which at least one of the toolchains completes in under 1000 seconds.

>>>>

Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
delete the input clauses.  Consider the first of these, deleting
input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
that there is no total assignment $\assign$ that falsifies this clause but assigns $\assign(s_{10}) = 1$.
The proof proceeds by first assuming that the clause is false, requiring
$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consist of unit
clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
clauses that cause unit propagation.  Hint clauses 8 and 10 force the
assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 force
$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading,
via clause 22, to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
completing the RUP proof.  The deletion hints for the other input
clauses follow similar patterns---they work from the bottom nodes of
the POG upward, showing that any total assignment that falsifies the clause
must assign $\assign(s_{10}) = 0$.

@@

Our experiments are designed to evaluate the following:
\begin{itemize}
\item
whether our toolchain can handle challenging benchmark problems,
\item
the effectiveness of some design choices and optimizations,
\item
whether our toolchain can perform end-to-end verification with preprocessed formulas,
\item
the comparative performance of the prototype and verified checkers, and
\item
how the performance of our toolchain compares to that of related verification toolchains.
\end{itemize}

@@

Our default configuration for the proof generator used the structural
method for the forward implication proof, with the two optimizations
of literal grouping and lemmas.
Running with a time limit of 4000 seconds,\footnote{All measured times
in this document are actual elapsed times, not CPU times.} \dfour{}
was able to generate dec-DNNF representations for 124 of these, and it timed out on the
rest.  Our proof generator was able to convert all of the generated
dec-DNNF graphs into POGs and determine how many defining clauses they
would generate.  A POG operation with $k$ arguments requires $k+1$
defining clauses, and so the total number of defining clauses in POG $P$ equals
the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$.
The number of defining clauses ranged from 304 to 2,761,457,765 with a median of 848,784.

@@

Figure~\ref{fig:d4:cpog} summarizes our results in terms of the time
required by \dfour{} (X axis) versus the sum of the times for proof generation, checking, and counting
 (Y axis).
We were able to complete a full validation for 108 of the 124
benchmarks, with times ranging from fractions of a second to 13,144
seconds, with a median of 71.6 seconds.

Relative to the runtime for \dfour{}, two problems ran faster with the
generator, checker, and counter.  These were for problems having small numbers
of models (relative to the number of variables), and so most of the
time spent by both \dfour{} and our proof generator was in running a
CDCL solver to search the very sparse solution space.  \cadical{}
generally outperforms the miniSAT-based solver used by \dfour{}.  At
the other extreme, one problem that required only 19 seconds for
\dfour{} required 8657 seconds to generate a proof and 45 seconds to
check it (and count).  This benchmark appears to stem from weak performance
by our implementation of BCP\@.  Overall the ratios
between the combined time for generation, checking, and counting versus the time
for \dfour{} had a harmonic mean of 5.5.

@@

To assess the relative merits of the two approaches to forward
implication proof generation, we ran the proof generator in monolithic
mode on the 92 problems for which the combination of structural proof generation, checking, and counting
required at most 3000 seconds and then kept only those
measurements for which at least one of the two approaches had combined
times of at most 1000 seconds.  There were
a total of 83 such problems.  Figures~\ref{fig:monolithic:time} and
\ref{fig:monolithic:clauses} show comparisons between the two
approaches in terms of time and total clauses.

Examining Figure~\ref{fig:monolithic:time}, we can see that the
monolithic approach ran faster than the structural approach for 23 of
the problems, including a case where the generator, checker, and counter
required only one second, versus 71 seconds for the structural
approach.  On the other hand, the monolithic approach timed out for 13
of the cases and was slower for the other 37.  In general, we can see
the monolithic approach faring worse on larger problems.

@@

As the figures illustrated, each optimization on its own can shorten
the proofs, sometimes substantially, and the two combine to have even
greater effect.  Of the 80 benchmark problems evaluated, and comparing
with both optimizations disabled, 47 had the number of proof clauses
reduced by at least a factor of 2 by using lemmas, 14 by using literal merging,
and 60 by enabling both optimizations.  In the extreme cases, a lack
of lemmas caused one proof to grow by a factor of 52.5, while a lack
of literal grouping caused another proof to grow by a factor of 39.6.

@@

One strength of monolithic proof generation is that it makes no
assumptions about how the compiled representation was generated.  A complete
proof-generating SAT solver could, in principle, always generate the
forward implication proof, as long as forward implication holds.  To
test this hypothesis, we ran our proof framework with preprocessing
enabled on the same 83 benchmarks as in
Section~\ref{app:experiment:monolithic}.  We measured the time for two
entire toolchains: one with compilation, proof generation, 
checking, and counting, and the other starting with preprocessing and then having
the same steps as the other.  The results are summarized in
Figures~\ref{fig:preprocess:time} and \ref{fig:preprocess:clauses}.
Setting a time limit of 1000 seconds, 70 of the formulas completed
without preprocessing, while 69 did so with preprocessing.  The one
that timed out with preprocessing did so due to the excessive time
required for preprocessing.  Of the 69, 53 ran faster with
preprocessing, while 15 ran slower.  The sizes of the generated 
CPOG proofs were more variable: 31 had fewer clauses with preprocessing, 21 had more, and 17 were identical.

@@

This section provides more insights into how our proof framework (CPOG)
compares with other work in certified knowledge compilation and model counting.

@@

To compare the performance of our toolchain versus that of
MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
generation, proof checking, and counting in the CPOG toolchain compares to the time to
run their proof generator \progname{nnf2trace} and checker
\progname{sharptrace}.
We ran their tools on the 92 benchmark problems for
which our toolchain requires at most 3000 seconds and retained
the 84 data points for which at least one of the toolchains completes in under 1000 seconds.
